Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-22 manuscript/climate_sensitivity_draft.Rmd
worked on #65

A final central question regarding forest responses to drought is whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. While the tendency for larger trees to experience more severe drought certainly predominates the literature [@bennett_larger_2015], there are exceptions [*e.g.*, @suarez_factors_2004; @zang_size-dependent_2012; @mcdowell_mechanisms_2008]. There is also evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.

@@

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected at breast heigh (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  **Ryan, have you heard back from ITRDB?  We should also cite GitHub/Zenodo here. **

@@

Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.

Crown positions ($CP$) were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible). An analysis of crown position relative to height (Fig. 2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions, although dominant and co-dominant trees were similar in height (Fig. S3). Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community. 
Candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth. 
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest. **see comment 25**

@@

For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011]. Analyses focused on testing the predictions presented in Table 1, with either $Rt$ or $H$ as the response variable. **see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.

@@

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

>>>>

A final central question regarding forest responses to drought is whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.

@@

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  **Ryan, have you heard back from ITRDB?  We should also cite GitHub/Zenodo here. **

@@

Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).**Ian, I think we need to include a little more about the NEON data here.** Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.

Crown positions ($CP$) were recorded for all cored trees that remained standing during the growing season of 2018. We followed the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible). An analysis of crown position relative to height (Fig. 2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions, although dominant and co-dominant trees were similar in height (Fig. S3). Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community. 
Candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth. 
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

@@

For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$<1 indicates reduced growth under drought. Analyses focused on testing the predictions presented in Table 1, with either $Rt$ or $H$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.

@@

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volume provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
